Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2025-01-15 demystify.tex
Updates from Overleaf

Moreover, depending on the university, you may receive additional benefits such as summer pay, laptops, and traveling to conferences and workshops. 
Full funding for CS PhD students is the norm in the US, and I'd go as far as to say that if you are not admitted with full funding, you might want to not accept the offer.

@@

Students from top schools with strong research programs and experience might already know most of the information given here. They might also receive guidance from their professors or grad students who have gone through the process. Students from smaller schools or countries, however, might not have access to such resources and might not know where to start. 
My goal is thus to level the playing field by providing info that is not readily available to less privileged students. 
I hope to encourage more students with such backgrounds to apply and succeed in CS PhD programs in the US.

@@

While this handbook may not fully meet the needs of students aiming for top-10 schools (\autoref{sec:selecting-ranking-schools} and \autoref{sec:ranking}), which typically have more stringent requirements (e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU), it should be valuable for those targeting other reputable R1 institutions.

@@

\begin{center}
  \includegraphics[scale=0.5]{files/c4a.png}
\end{center}

@@

The PhDComics figure above shows the ``ambition'' level of a PhD student over their years of study (they miss the 6--7th year when the ambition is \emph{``Just let me graduate''}).

@@

A main question adcom has for a non-CS or non-STEM student is if you have the sufficient technical background obtained through core CS courses.  So you need to show that you have such knowledge through your coursework, projects, or research.
For example, if you have taken a course on Algorithms, even online ones like Coursera, you can talk about it in your SOP.  If you have done a project that requires knowledge of OS or have a professional certification (e.g., A+) through work, you can talk about it.  If you have done research that requires knowledge of Discrete Maths, you can talk about it.  You can also ask your LoR writers to talk about your technical background.
In summary, in your application, convince us that you have the background to do CS PhD research.

@@

No, while other countries often require an MS for PhD student in CS (~\autoref{sec:non-us-differences}), it is common in he US to apply for a PhD program directly after a 4-year undergrad program (e.g., after getting a B.S degree).  Most CS PhD programs are designed so that students can get MS degree ``along the way'' to PhD, e.g., after finishing the 2-year course work.

@@

\emph{Course work}: In the US ou will spend the first couple of years taking classes and exploring potential adviser and research topics. 
After that, you have to pass a series of exams during your PhD, e.g., qualifying exam, comprehensive exam, thesis proposal defense\footnote{The word ``ABD'' (all but dissertation) is used in the US to refer to a PhD candidate who has finished all course work and exams and only needs to write and defend their dissertation.}.

@@

\section{Admission Committee (adcom)}\label{sec:adcom}
\sectioninfo{Adcom members are faculty who evaluate your application. They consider various factors, e.g., research experience, LoRs, SOP}

@@

Each application is assigned to about three adcom members, who will evaluate your profile and reach a consensus.  They will consider various factors, e.g., research experience, GPA, LoRs, SOP, and test scores. More details about these factors are discussed in \autoref{part:appendices}.

\begin{commentbox}{Vu:}
At Mason, we usually decide that a full-time PhD candidate is either (i) admitted with funding (\autoref{sec:funding}) or (ii) rejected. In other words, in most cases, we either
admit you with full funding or reject your application. In some rare cases, we admit
without funding because you have funding on your own, e.g.,
supported by your government or having external fellowships. We justify
our decision (\autoref{sec:ievaluate}) with a summary of your application, where we list
strengths, e.g., a well-known school, and weaknesses, e.g., weak LoRs.
\end{commentbox}

@@

Not really, we typically review applications independently and do not talk to each other.  This is to avoid biasing, e.g., if one reviewer says they want to accept, the other might feel pressured to accept as well.

@@

\begin{commentbox}[A prof. in Physics at an R1 university:]
  We have a pretty well fleshed out grading rubric for applications that has categories like grades, research, writing ability, etc. I would say our rubric is weighted about 1/2 on academics (research, LORs, grades) and 1/2 on the idea of ``grit'' or ``resilience'' (engagement, leadership, working through obstacles).\\

@@

\section{Why CS depts do not waive the application fee?}\label{sec:fee-waive}

Some universities do waive, e.g., Rice, TTIC do not have fee for PhD applications and many universities for domestic students (\autoref{sec:domestic-students}), however most do not for international students.
Some require proof of financial difficulties, e.g., a statement from a financial adviser or a bank statement.

In general, application fee is typically a requirement of the university. Individual departments and programs do not have the flexibility to waive the application fee, even if they want to.

\begin{commentbox}{Vu:}
In my opinion, requiring applicants to pay the fee helps ensure their seriousness, as it filters out non-serious candidates. Most CS programs already receive many applications and would be overwhelmed if the application process were free.  Even with application fee the competition is already very tough, imagine if the application is free and the number of applications doubles or triples.\\

@@

Most PhD programs will require at least \emph{two} letter of recommendations (LoRs). 
The goal of the LoRs is to provide the adcom with an assessment of your research ability from someone who knows you.  The LoRs should focus on your research background, achievements, and potentials, all of which paint a compelling picture of you as a researcher to adcom.

@@

\subsection{LoR from Famous People}  Having a strong letter from an internationally recognized researcher can \emph{greatly strengthen} your application. Such a letter can outweigh the lack of publications (or low GPA or unknown school). A well-known researcher has a reputation to maintain (so they won't recommend someone bad), and adcom member trust people that they know, or heard about and respect (compared to someone they don't know).

>>>>

Moreover, you often receive additional benefits such as summer pay (\autoref{sec:summer-funding}), laptops (\autoref{sec:buying-equipment}), and traveling to conferences and workshops. 
Full funding for CS PhD students is the norm in the US, and I'd go as far as to say that if you are not admitted with full funding, you might want to not accept the offer.

@@

Students from top schools with strong research programs and experience might already know most of the information given here. They might also receive guidance from their professors or grad students who have gone through the process. Students from smaller schools or countries, however, might not have access to such resources and know where to start. 
My goal is thus to level the playing field by providing info that is not readily available to less privileged students. 
I hope to encourage more students with such backgrounds to apply and succeed in CS PhD programs in the US.

@@

%While this handbook may not fully meet the needs of students aiming for top-10 schools (\autoref{sec:selecting-ranking-schools} and \autoref{sec:ranking}), which typically have more stringent requirements (e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU), it should be valuable for those targeting other reputable R1 institutions.

@@

\begin{figure}
\centering
  \includegraphics[scale=0.5]{files/c4a.png}
  \caption{The ``ambition'' level of a PhD student over their years of study (they miss the 6--7th year when the ambition is \emph{``Just let me graduate''}).}
\end{figure}

@@

@@

A main question adcom (\autoref{sec:adcom}) has for a non-CS or non-STEM student is if you have the sufficient technical background obtained through core CS courses.  So you need to show that you have such knowledge through your coursework, projects, or research.
For example, if you have taken a course on Algorithms, even online ones like Coursera, you can talk about it in your SOP.  If you have done a project that requires knowledge of OS or have a professional certification (e.g., A+) through work, you can talk about it.  If you have done research that requires knowledge of Discrete Maths, you can talk about it.  You can also ask your LoR writers to talk about your technical background.
In summary, in your application, convince us that you have the background to do CS PhD research.

@@

No, while other countries often require an MS for PhD student in CS (\autoref{sec:non-us-differences}), it is common in the US to apply for a PhD program directly after a 4-year undergrad program (e.g., after getting a B.S degree).  Most CS PhD programs are designed so that students can get MS degree ``along the way'' to PhD, e.g., after finishing the 2-year course work.

@@

\emph{Course work}: In the US ou will spend the first couple of years taking classes and exploring potential adviser and research topics. 
After that, you have to pass a series of exams during your PhD, e.g., qualifying exam, comprehensive exam, thesis proposal defense\footnote{ABD (all but dissertation) refers to a PhD candidate who has finished all course work and exams and only needs to write and defend their dissertation.}.

@@

\section{Admission Committee (Adcom)}\label{sec:adcom}
\sectioninfo{Adcom members are faculty who evaluate your application. They consider various factors, e.g., research experience, LoRs, SOP}

@@

Each application is assigned to about three adcom members, who will evaluate your profile and reach a consensus.  They will consider various factors, e.g., research experience, GPA, LoRs, SOP, test scores, and interviews. More details about these factors are discussed in \autoref{part:appendices}.

\begin{commentbox}{Vu:}
At Mason, we usually decide that a full-time PhD candidate is either (i) admitted with funding (\autoref{sec:funding}) or (ii) rejected. In other words, in most cases, we either
admit you with full funding or reject your application. In some rare cases, we admit
without funding because you have funding on your own, e.g.,
supported by your government or having external fellowships. We justify
our decision (\autoref{sec:ievaluate}) with a summary of your application, where we list
strengths, e.g., came from a well-known school, and weaknesses, e.g., weak and generic LoRs.
\end{commentbox}

@@

We typically review applications independently and do not talk to each other.  This is to avoid biasing, e.g., if one reviewer says they want to accept, the other might feel pressured to accept as well.

@@

\begin{commentbox}[From a prof. in Physics at an R1 university:]
  We have a pretty well fleshed out grading rubric for applications that has categories like grades, research, writing ability, etc. I would say our rubric is weighted about 1/2 on academics (research, LORs, grades) and 1/2 on the idea of ``grit'' or ``resilience'' (engagement, leadership, working through obstacles).\\

@@

\section{Waiving the application fee}\label{sec:fee-waive}

Some universities do waive, e.g., Rice, TTIC do not have fee for PhD applications and many universities for domestic students (\autoref{sec:domestic-students}).
Some programs also waive if the applicants attend some of their opening sessions.  Some programs waive if the applicant provide proof of financial difficulties, e.g., a statement from a financial adviser or a bank statement.

However most do not waive the application fee, which is typically a requirement of the university. Individual departments and programs do not have the flexibility to waive the application fee, even if they want to.

\begin{commentbox}{Vu:}
In my opinion, requiring applicants to pay the fee helps ensure their seriousness, as it filters out non-serious candidates. Most CS programs already receive many applications and would be overwhelmed if the application process were free.  Even with application fee the competition is already very tough, imagine if the application is free and the number of applications triples or quadruples.\\

@@

Most PhD programs will require at least \emph{two} letter of recommendations (LoRs). 
The goal of the LoRs is to provide the adcom with an assessment of your research ability from someone who knows about research and \emph{you} well.  The LoRs should focus on your research background, achievements, and potentials, all of which paint a compelling picture of you as a researcher to adcom.

@@

\subsection{LoR from Well-Known People}  Having a strong letter from an internationally recognized researcher can \emph{greatly strengthen} your application. Such a letter can outweigh the lack of publications (or low GPA or unknown school). A well-known researcher has a reputation to maintain (so they won't recommend someone bad), and adcom member trust people that they know, or heard about and respect (compared to someone they don't know).
